---
title: "Complete Data Analysis Workflow in R"
author: "By Muhammad Yasir Qurashi"
date: "`r Sys.Date()`"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## **Step 01**

> Defining Research Question

Investigate morphological variation among penguin species and examine relationships between body size measurements, and answer the following queries;

-   Do penguin species differ in body mass?

-   Is flipper length associated with body mass?

-   Are there morphological differences between male and female penguins?

## **Step 02**

> Load Require Libraries

```{r}
# Keep remember first install these libraries in your console and then activate using library command

library(tidyverse)
library(palmerpenguins)
library(naniar)
library(patchwork)
library(GGally)
library(janitor)
library(skimr)
library(rstatix)
```

| Package   | Role                       |
|-----------|----------------------------|
| tidyverse | data manipulation          |
| janitor   | clean column names         |
| skimr     | dataset summary            |
| naniar    | missing data visualization |
| GGally    | correlation plots          |
| patchwork | combine figures            |
| rstatix   | check outliers in once     |

## **Step 03**

> Load dataset

```{r}
penguin_data <- penguins
print(penguin_data)
```

## **Step 04**

> Understanding your data

```{r}
dim(penguin_data)
nrow(penguin_data)
ncol(penguin_data)
colnames(penguin_data)
head(penguin_data)
str(penguin_data)
glimpse(penguin_data)
tail(penguin_data)
slice_sample(penguin_data)
summary(penguin_data)
```


| Variable          | Type        |
|-------------------|-------------|
| species           | categorical |
| island            | categorical |
| bill_length_mm    | numeric     |
| bill_depth_mm     | numeric     |
| flipper_length_mm | numeric     |
| body_mass_g       | numeric     |
| sex               | categorical |
| year              | numeric     |

-   No of rows(observation); 344
-   No of columns(variables); 8

## **Step 05**

> Data Quality assessment 

```{r}
# 1. Check duplicates

duplicated(penguin_data)
sum(duplicated(penguin_data))

## No duplicate values detected

# 2. Summary statistics to check unusual values

skim(penguin_data)

```

## **Step 06**

> Data Cleaning


```{r}
# 1. Standardize column names
penguin_data <- penguin_data %>% clean_names() 
 colnames(penguin_data)
glimpse(penguin_data)
```

## **Step 07**

> Handling missing values

```{r}
# 1. Checking for missing values in each column

is.na.data.frame(penguin_data)
is.na(penguin_data)
sum(is.na(penguin_data)) # total number of missing values in a dataframe
colSums(is.na(penguin_data)) # column wise missing values of our dataset
vis_miss(penguin_data)
penguin_data %>% is.na() %>% colSums() %>% barplot() # plottting missing values in our dataset

# 2. remove missing values rows in each column

penguin_data <- penguin_data %>% drop_na()

dim(penguin_data) # 11 rows with missing values are removed from our dataset

penguin_data %>% is.na() %>% colSums() %>% barplot() # Now their is no missing values in each column
```

## **Step 08**

> Outlier Detection

```{r}
# 1. Checking outliers by rstatix library

penguin_data %>% 
  select(where(is.numeric)) %>% 
  gather(variable, value) %>%
  group_by(variable) %>%
  identify_outliers(value)

# No outliers detected

# 2. Checking outliers by boxplot

pg <- penguin_data %>%  
  select(where(is.numeric)) %>% 
  select(-year) %>% 
  pivot_longer(everything(), names_to = "variable", values_to = "value") %>% 
  ggplot(aes(x = variable, y = value)) +
  geom_boxplot() +
  facet_wrap(~variable, scales = "free_y");pg

# No outliers detected
```

## **Step 09**

> Data transformation

If transformation is necessary we will do it, in our case we do for body_mass variable


```{r}

penguin_data1 <- penguin_data %>% 
  mutate(body_mass_kg = body_mass_g / 1000)
ncol(penguin_data1)
summary(penguin_data1)
```

## **Step 10**

> Exploratory Data Analysis

Exploring patterns in our data

```{r}
# 1. Histogram

ggplot(penguin_data, aes(body_mass_g)) +
   geom_density() ; ggplot(penguin_data, aes(flipper_length_mm)) +
   geom_density() ; ggplot(penguin_data, aes(bill_depth_mm)) +
   geom_density() ; ggplot(penguin_data, aes(bill_length_mm)) +
   geom_density()
```

```{r}
# 2. Boxplot

ggplot(penguin_data, aes(species, body_mass_g)) +
  geom_boxplot() ; ggplot(penguin_data, aes(species, flipper_length_mm)) +
  geom_boxplot() ; ggplot(penguin_data, aes(species, bill_length_mm)) +
  geom_boxplot() ; ggplot(penguin_data, aes(species, bill_depth_mm)) +
  geom_boxplot()

```

```{r}
# 3. Correlation

ggpairs( penguin_data1, columns = c( "bill_length_mm","bill_depth_mm","flipper_length_mm","body_mass_kg"), 
         lower = list(continuous = wrap("points", alpha = 0.6, size = 2)),
         upper = list(continuous = wrap("cor", size = 5)),
         diag = list(continuous = wrap("densityDiag")))
```


## **Step 10**

> Statistical test & Regression Analysis

Test body mass differences between species  

```{r}
 anova_model <- aov(body_mass_g ~ species, data = penguin_data)
 summary(anova_model)

# Conclusion
 
# Significant species differences

```

```{r}
  
model <- lm(body_mass_g ~ flipper_length_mm, data = penguin_data)
 summary(model)
 
# Conclusion
 
# Flipper length significantly predicts body mass. 
```


## **Step 11**

> Publication Ready Visualization

```{r}
ggplot(penguin_data, aes(x = flipper_length_mm, y = body_mass_g, color = species)) +
  geom_point(size = 3) +
  geom_smooth(method = "lm") + 
  theme_classic()
```

> Intepretation of Analysis

Penguin species differed significantly in body mass (ANOVA, p < 0.001). Gentoo penguins exhibited the highest body mass, whereas Adelie penguins were the smallest. A strong positive association between flipper length and body mass was also observed, suggesting that larger penguins possess proportionally longer flippers


Best Regards,

Muhammad Yasir Qurashi

*Research Data Analysis Tools Mentor*
